---
title: "Categorical Auto-RQA"
author: "Fred Hasselman"
date: "1/14/2019"
output: 
  bookdown::html_document2: 
    variant: markdown+hard_line_breaks
    fig_caption: yes
    highlight: pygments
    keep_md: yes
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: true
    collapsed: false
    smooth_scroll: true
    code_folding: show
    bibliography: [refs.bib, packages.bib]
    biblio-style: apalike
    csl: apa.csl
    includes:
        before_body: assignmentstyle.html
    pandoc_args: ["--number-offset","4"]
editor_options: 
  chunk_output_type: console
---


```{r, include=FALSE}
knitr::opts_chunk$set(include=TRUE, tidy = FALSE)
```


# (Cross-) Recurrence Quantification Analysis Software

There are several packages in `R` you could use to do `(C)RQA` analyses, but you can also find a `Matlab` toolbox for `(C)RQA` here: [CRP toolbox](http://tocsy.pik-potsdam.de/CRPtoolbox/), there are `Python` libraries as well, see [this page](http://www.recurrence-plot.tk/programmes.php) for an overview of software. We will use the functions in package `casnet`.

```{block2, crqa, type='rmdimportant'}
 The package `crqa()` was mainly designed to run categorical Cross-Recurrence Quantification Analysis (see [Coco & Dale (2014)](http://journal.frontiersin.org/Journal/10.3389/fpsyg.2014.00510/abstract) and for R code see appendices in [Coco & Dale (2013)](http://arxiv.org/abs/1310.0201)). The article is a good reference work for details on how to conduct RQA analysis.
```


We'll use the`casnet`:

```{r, eval=FALSE}
# Install casnet if necessary: https://github.com/FredHasselman/casnet
# !!Warning!! Very Beta...
# library(devtools)
# devtools::install_github("FredHasselman/casnet")

library(casnet)
library(rio)
```


Package [`casnet`](https://fredhasselman.github.io/casnet) has 2 functions that will perform auto- and cross-recurrence quantification analyses:         

* Function `rp_cl()` uses a precompiled program that is run from the command line (see [command line recurrence plot tools](http://tocsy.pik-potsdam.de/commandline-rp.php)). It will take as input one or two timeseries and construct a recurrence matrix and is very fast compared to other methods. The first time you run the command `rp_cl()` the `casnet` package will try to download the correct executables for your operating system, look for messages in the console to find out if the script succeeded. The output of `rp_cl()` will be similar to the [CRP toolbox](http://tocsy.pik-potsdam.de/CRPtoolbox/) and [`Python` libraries](http://www.recurrence-plot.tk/programmes.php) by Norbert Marwan.
  - **This function will no longer work on MacOS Catalina and later**, because, it is a 32 bit program, use `rp()` and `rp_measures()`
  - If downloading or running `rp_cl()` fails on OS that allows 32 bit programs, this is usually because you do not have enough rights to execute a command line program on your machine. In that case, figure out how to get those rights, or use `rp()` and `rp_measures()`.
* Function `rp_measures()` assumes you have already contructed a recurrence matrix using function `rp()`. The output of `rp_measures()` contains more measures than `rp_cl()` some of which are still experimental.


## **Categorical Auto-RQA** 


### RQA by 'hand' {.tabset .tabset-fade .tabset-pills}

Let's start with a small time series... open and/or download this Google sheet: https://docs.google.com/spreadsheets/d/1k4FYDbszfCReCM5_wcw6NIEgSFtTPIy3qmf-kFiZjy8/edit?usp=sharing 

The spreadsheet displays a short categorical time series of 5 observed categories labelled `A` through `E`.
Below the time series a matrix is displayed with zeroes in the upper triangle. The columns and rows of the matrix are labelled according to the categorical time series **y** in `B2:B13`, starting in the lower left corner. The *Line Of Incidence* (LOI) is the diagonal of the matrix going from the lower left corner to the upper right corner. Like the diagonal of a correlation matrix, the LOI in Auto-RQA will contain only recurrent points (`1`), because that's where the column labels will be identical to the row labels.


#### Questions {-}

Because this is Auto-RQA and we are comparing the time series to itself, the recurrence matrix will be symmetrical, so we only have to look at the upper triangle.

* Fill the upper triangle with recurrent points by changing a `0` into a `1`
  - Start with the label in cell `C29` (**A**) and look upwards (`C27` to `C17`). If you see the value **A** recurring, change the `0` into a `1`
   - Proceed to `D29` and find recurrences of the value **B** in the future (in `D26` to `D17`)
   - Continue to fill the upper triangle
* Count the number of `1`s, the recurrent points (`RN`) and put the value in cell `B32` (*Total*)
* Count how many recurrent points are part of a vertical line, and how many are on a diagonal line, put the values in cell `B33` and `B34` respectively.
* Diagonal line lengths:
  - Determine the frequency of occurrence of specific lengths of diagonal lines, record the values in cells `Q17` to `Q20`
  - Record the maximal diagonal line length in `V22`
  - The mean of the lengths of diagonal lines (*L*) will be calculated in cell `V21`
* Vertical line lengths: 
  - Determine the frequency of occurrence of specific lengths of vertical lines, record the values in cells `Q24` to `Q27`
  - Record the maximal vertical line length in `V27`
  - The mean of the lengths of vertical lines (*Trapping Time*) will be calculated in cell `V21`
* Several values are auto-calculated, if you click on the cell you can see how this is done, by looking at the formula bar.
  - Can you understand the calculation of the number of possible recurrent points in `X17`?
* Suppose you randomise the time series... which of the following values do you think would change, and in what way (larger/smaller)
  - Recurrence Rate (`RR`)
  - Mean diagonal line length (`L`, or `MEAN_dl`)
  - Maximum diagonal line length (`L_max`, or `MAX_dl`)
  - Entropy of distribution of diagonal line lengths (`L_entr`, or `ENT_dl`)

#### Answers {-}

* The correct matrix is on the second tab of the spreadsheet.
* The calculation of the number of possible recurrent points in cell `X17` is the size of the matrix (12*12) divided by 2 (because we are only looking at the upper triangle), minus the length of the diagonal, because these aren't actually recurrent points.
* Which of the following values do you think would change if `y` is randomised?
  - Recurrence Rate (`RR`) - **No change**, none of the values are changed, so the number of values that will recur is the unchanged.
  - Mean diagonal line length (`L`, or `MEAN_dl`) - **Likely to be lower**, shuffling will likely break up line structures, but if there are just a few categories and/or the time series is short, it could be the same or higher.
  - Maximum diagonal line length (`L_max`, or `MAX_dl`) - **Very likely to be lower**, shuffling will likely break up line structures, but if there are just a few categories and/or the time series is short, it could be the same or higher.
  - Entropy of distribution of diagonal line lengths (`L_entr`, or `ENT_dl`) - - **Difficult to predict**, shuffling will break up line structures and could increase entropy (more different line lengths), but it is also possible that ther will be only lines of length 2 or 3 left after randomisation, this will be a very homogeneous distribution which has a high entropy. This is more likely if there are just a few categories and/or the time series is short.


### RQA and executive control

`casnet` comes with a vignette demonstrating the usage of `rp_cl()`, see https://fredhasselman.github.io/casnet/articles/cl_RQA.html, or the manual pages. 

* Study the vignette and try to understand the conclusion based on the surrogate tests.


# Phase Space Reconstruction and Continuous Auto-RQA


## **Phase Space Reconstruction** 


Here is a great video summary of the phase space reconstruction technique: https://youtu.be/6i57udsPKms


```{block2, type='rmdnote'}
**R-packages for Phase Space Reconstruction**

Package `casnet` relies on packages `fractal` and `rgl` to reconstruct phase space. Package `rgl` is used for plotting and can cause problems on some windows machines. On windows You'll need to check if you have the [X Window System](http://www.x.org/wiki/) for interactive 3D plotting. This Linux desktop system comes installed in some way or form on most Mac and Windows systems. You can test if it is present by running `rgl::open3d()` in `R`, which will try to open an interactive plotting device.

If this fails you can try package `plot3D` which is not interactive function 
```


### Reconstruct the Lorenz attractor {.tabset .tabset-fade .tabset-pills}

Package `fractal` includes the 3 dimensions of the Lorenz system in the chaotic regime. Run this code `rgl::plot3d(lorenz,type="l")` with all three packages loaded to get an interactive 3D plot of this strange attractor.

We'll reconstruct the attractor based on just dimension `X` of the system using the data from package `fractal`, and functions from package `casnet`. Don't forget: **always look at the manual pages of a function** you are running.

```{block2, samepack, type='rmdimportant'}
Package `fractal` and package `nonlinearTseries` use functions with similar names, perhapp better to not load them together, or always use `fractal::` and `nonlinearTseries::` if you want to call the functions direclty.
```


#### Questions {-}

* Use `lx <- lorenz[1:1024,1]` to reconstruct the phase space based on `lx`. 
    + Find an optimal embedding lag and dimension by calling `est_parameters()`.
    + Use `nnThres = .01`. This is the proportion of remaining nearest neighbours below which we consider the number of dimensions optimal. Because this time series was generated by a simulation of a system, we can use a much lower threshold. For real data, which is much noisier, you can use `0.1` or `0.2`.
    + **NOTE:** If you have a slow computer don't use `1024`, but use e.g. `512`.
    
This `est_parameters()` function will call several other functions (e.g. `fractal::timeLag()` with `method = "mutual"` and `nonlinearTseries::findAllNeighbours()` for the false nearest neighbourhood search). Look at the manual page, you can run the `est_parameters()` function with the defaults settings, it will show a diagnostic plot. Because we have to search over a range of parameters it can take a while to get the results...

* Based on the output:
    + Choose an embedding lag $\tau$ (in `casnet` functions this is the argument `emLag`). Remember, this is just an optimisation choice. Sometimes there is no first minimum, or no global minimum.
    + Find an appropriate embedding dimension by looking when the false nearest neighbours drop below 1\%. 
    
* Now you can use this information to reconstruct the phase space: 
    + Embed the time series using `ts_embed()` (you can also use `fractal::embedSeries()`).
    + Inspect the object that was returned
    + Use  `rgl::plot3d()` to plot the reconstructed space. Plot the reconstructed phase space. (You'll need to use `as.matrix()` on the object created by `fractal::embedSeries()`)

#### Answers {-}

* Use `lx <- lorenz[1:1024,1]` to reconstruct the phase space based on `lx`. 
    + Find an optimal embedding lag and dimension by calling `est_parameters()`.

```{r , cache=TRUE}
library(rgl)
library(ggplot2)
library(fractal)
library(invctr)
library(casnet)

# The X data
lx <- lorenz[1:1024,1]

# Search for the parameters,
params <- est_parameters(lx)

# Assign the optimal parameters
emLag <- params$optimLag
emDim <- params$optimDim
```

* Now you can use this information to reconstruct the phase space

```{r, eval = TRUE, cache=TRUE}
# For X
lx_emb <- ts_embed(y=lx,emLag = emLag, emDim = emDim)

rgl::plot3d(lx_emb, type="l")

# For Y
ly <- lorenz[1:1024,2]

# Search for the parameters,
params <- est_parameters(ly)

# Assign the optimal parameters
emLag <- params$optimLag
emDim <- params$optimDim

# Embed the time series
ly_emb <- ts_embed(y=ly,emLag = emLag, emDim = emDim)
rgl::plot3d(ly_emb,type="l")

# For Z
lz <- lorenz[1:1024,3]

# Search for the parameters,
params <- est_parameters(lz)

# Assign the optimal parameters
emLag <- params$optimLag
emDim <- params$optimDim

# Embed the time series
lz_emb <- ts_embed(y=lz,emLag = emLag, emDim = emDim)
rgl::plot3d(lz_emb,type="l")
```


## Continuous Auto-RQA


### RQA of the Lorenz attractor {.tabset .tabset-fade .tabset-pills}

Perform an RQA on the reconstructed state space of the Lorenz system. You'll need a radius (also called: threshold, or $\epsilon$) in order to decide which points are close together (recurrent). You have already seen `casnet` provides a function which will automatically select the best parameter settings for phase space reconstruction: `est_parameters()`. 


#### Questions {-}

* First get a radius that gives you a fixed recurrence rate use `est_radius()` look at the manual pages. The function `rp()` will also call `est_radius` if you provide `emRad = NA`.

* Best way to ensure you are using the same parameters in each function is to create some lists with parameter settings (check the `rp_cl()` or `rp()` manual to figure out what these parameters mean)


#### Answers{-}

```{r, cache=TRUE}
library(fractal)
library(casnet)

# Lorenz X
lx <- lorenz[1:1024,1]
emLag = 17
emDim = 3

(emRad <- est_radius(y1 = lx, emLag = emLag, emDim = emDim)$Radius)

# RQA analysis
#(out <- rp_cl(lx,emDim = emDim, emLag = emLag, emRad = emRad))
RMx <- rp(y1 = lx, emDim = emDim, emLag = emLag, emRad = emRad)
out <- rp_measures(RMx)
```


We can plot the recurrence matrix...

```{r}
library(casnet)

# Unthresholded matrix (no radius applied)
RM <- rp(y1 = lx, emDim = emDim, emLag = emLag)

# plot it
rp_plot(RM)


# Thresholded by the radius
RMth <- di2bi(RM,emRad = emRad)

# You cam also call rp wth emRad = NA
RMth <- rp(y1 = lx, emDim = emDim, emLag = emLag, emRad = NA)


rp_plot(RMth, plotDimensions = TRUE, plotMeasures = TRUE)

```


### Distinguish between Logistic Map and White noise? {.tabset .tabset-fade .tabset-pills}

RQA promises to be able to distinguish between different dynamical regimes, for example dynamics due to different parameter settings of the logistic map, see e.g. https://en.wikipedia.org/wiki/Recurrence_quantification_analysis#Time-dependent_RQA 


#### Questions {-}

* Use RQA to compare white noise (TS_1) to the deterministic chaos series of previous assignments (TS_3).
   + You can shorten the time series by half to run RQA on a slow computer.
* Estimate the parameters
   + Did you expect to get these values?
   + Choose the same lag and dimension for both series, use the median of the lags and the maximum of the dimension.
* Produce an unthresholded matrix
* Choose a recurrence threshold to get `RR = 0.05` (either by looking at the color bar or by estimating with a function) and plot the recurrence plot.
* Compare the RQA measures.
   + Think about the properties of deterministic chaos when you interpret the results.
   + RQA is very powerful when detecting different dynamical regimes: 
https://en.wikipedia.org/wiki/Recurrence_quantification_analysis#Example



```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
# Reload the data
library(rio)
series <- rio::import("https://github.com/complexity-methods/CSA-assignments/raw/master/assignment_data/BasicTSA_arma/series.xlsx")

```

#### Answers {-}

* Shorten the series
```{r}
TS1 <- series$TS_1[1:512]
TS3 <- series$TS_3[1:512]
```


* Parameters
  + Remember that you need at least 3 coupled differential equations to get deterministic chaos.
  
```{r}
library(casnet)
out1 <- est_parameters(TS1) 
out3 <- est_parameters(TS3) 

emLag <- median(out1$optimLag,out3$optimLag)
emDim <- max(out1$optimDim,out3$optimDim)
```

* Plots
```{r}
# Unthresholded
RM1 <- rp(y1 = TS1, emDim = emDim, emLag = emLag)
rp_plot(RM1, plotDimensions = TRUE)

# Thresholded
RM1t <- rp(y1 = TS1, emDim = emDim, emLag = emLag, emRad = NA)
rp_plot(RM1t, plotDimensions = TRUE)

# Unthresholded
RM3 <- rp(y1 = TS3, emDim = emDim, emLag = emLag)
rp_plot(RM3, plotDimensions = TRUE)

# Thresholded
RM3t <- rp(y1 = TS3, emDim = emDim, emLag = emLag, emRad = NA)
rp_plot(RM3t, plotDimensions = TRUE)
```

* Measures 
```{r}
library(casnet)
rqa1 <- rp_measures(RM1t, silent = FALSE) 
rqa3 <- rp_measures(RM3t, silent = FALSE) 
```



### RQA of circle-tracing data {.tabset .tabset-fade .tabset-pills}

We'll analyse a time series recorded from a subject who is tracing a circle using a computer mouse:   [circle tracing data](https://github.com/FredHasselman/The-Complex-Systems-Approach-Book/tree/master/assignments/assignment_data/RQA_circletrace).


```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
# Reload the data
library(rio)
mouseXY <- rio::import("https://raw.githubusercontent.com/FredHasselman/The-Complex-Systems-Approach-Book/master/assignments/assignment_data/RQA_circletrace/mouse_circle_xy.csv")

```


#### Questions {-}

* Perform an RQA on one of the coordinate series (x or y)
* Study what happens to the RQA measures if you shuffle the temporal order.
* Package `fractal` contains a function `surrogate`. This will create so-called *constrained* realisations of the time series. You can look at the `rqa` vignette in package `casnet` ([also available on Github](https://fredhasselman.github.io/casnet/)) Look at the help pages of the function, or study the *Surrogates Manual* of the [TISEAN software](http://www.mpipks-dresden.mpg.de/~tisean/Tisean_3.0.1/index.html) and create two surrogate series, one based on `phase` and one on `aaft`.
     + Look at the RQA measures and think about which $H_0$ should probably be rejected.
     + If you want to be more certain, you'll have to create a test (more surrogates). The [TISEAN manual](http://www.mpipks-dresden.mpg.de/~tisean/Tisean_3.0.1/docs/surropaper/node5.html#SECTION00030000000000000000) provides all the info you need to construct such a test:
     
     > "For a minimal significance requirement of 95\% , we thus need at least 19 or 39 surrogate time series for one- and two-sided tests, respectively. The conditions for rank based tests with more samples can be easily worked out. Using more surrogates can increase the discrimination power."
  
#### Answers {-}

* will be discussed next week.


### EXTRA: Convergent cross mapping

Have a look at the package [`rEDM`, you can use it to do a convergent cross mapping analysis](https://cran.r-project.org/web/packages/rEDM/vignettes/rEDM-tutorial.html)



[Sugihara, G., May, R., Ye, H., Hsieh, C. H., Deyle, E., Fogarty, M., & Munch, S. (2012). Detecting causality in complex ecosystems. science, 338(6106), 496-500.](https://science.sciencemag.org/content/338/6106/496)


